---
title: "Reproducible Research - Assignment 1"
author: "Paul Lacock"
date: "27/12/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction
This is the output from the R Markdown produced for Assignment 1 of the Reproducible Research course.

## Loading and preprocessing the data
The source data has been downloaded from https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip and unzipped into the working directory.  

Load the data into a dataframe and convert the date field to the date class.
```{r data}
startdata <- read.csv("activity.csv")
startdata$date <- as.Date(startdata$date, format="%Y-%m-%d")
```
## Mean total steps per day
Calculate the total number of steps per day and plot as a histogram.
```{r totalsteps}
dailysteps <- as.vector(tapply(startdata$steps, startdata$date, FUN="sum", na.rm=TRUE))
hist(dailysteps, breaks = 24, main="Histogram of total daily steps", xlab="Total Daily Steps", ylab="Number of days")

# calculate the mean and median daily steps
meansteps <- mean(dailysteps, na.rm = TRUE)
mediansteps <- median(dailysteps, na.rm = TRUE)
meansteps
mediansteps
```
The mean number of steps per day is `r round(meansteps, 2)` and the median is `r as.integer(mediansteps)`.

## Daily activity pattern
The following time series plot shows the average number of steps taken per 5 minute time interval, over the 2 month observation period.
```{r activitypattern}
# derive mean steps per interval over the observation period and plot
intSteps <- aggregate(list(steps=startdata$steps), by = list(interval = startdata$interval), FUN = "mean", na.rm = TRUE)
plot(intSteps$interval, intSteps$steps, type = "l", main = "Daily average activity pattern", xlab = "Intervals", ylab = "Steps")

# identify the interval with the highest average steps
maxIntSteps <- max(intSteps$steps)
maxInt <- intSteps[intSteps$steps==maxIntSteps, ]
maxInt
```
The interval with the highest average number of steps is `r maxInt$interval` with an average of `r as.integer(maxInt$steps)` steps.

## Identifying and imputing missing values
Identify and count the rows with missing values.
```{r missing}
missing <- is.na(startdata$steps)
totmissing <- sum(missing)
totmissing
```
The number of rows with missing steps values is `r totmissing`.

A new dataset is created with the missing values imputed with the average number of (non-missing) steps per interval. The new dataset is then used to recalculate the total steps per day to produce a histogram which can be compared to the equivalent histogram based on the data without imputed values.
```{r impute}
newSteps <- startdata
newSteps$steps <- ifelse(missing, intSteps[newSteps$interval,]$steps, newSteps$steps)

# calc new daily totals including imputed values
NewDailyTot <- as.vector(tapply(newSteps$steps, newSteps$date, FUN="sum", na.rm=TRUE))

# plot the histogram
par(mfrow=c(2,1))
hist(dailysteps, breaks = 24, main="Histogram of total daily steps (missing)", xlab="Total Daily Steps", ylab="Number of days")
hist(NewDailyTot, breaks = 24, main="Histogram of total daily steps (imputed)", xlab="Total Daily Steps", ylab="Number of days")
newmeansteps <- mean(NewDailyTot, na.rm = TRUE)
newmediansteps <- median(NewDailyTot, na.rm = TRUE)
newmeansteps
newmediansteps
```
With imputed values the average number of daily steps has increased to `r round(newmeansteps, 2)` and the median is unchanged at `r as.integer(newmediansteps)`. Comparing the charts it is evident that the main effect of imputing has been to shift days showing very low total steps, and therefore not impacted the median.

## Activity patterns on weekdays and weekends
By introducing a new factor (based on date) we can compare the average activity patterns on weekdays and weekends.
```{r weekdays}
# create new factor
newSteps$weekday <- as.factor(ifelse(weekdays(newSteps$date) %in% c("Saturday","Sunday"), "Weekends", "Weekdays"))

# calculate average steps per interval, split between weekdays and weekends
weekdaySummary <- aggregate(list(steps=newSteps$steps), by = list(interval=newSteps$interval, weekday=newSteps$weekday), FUN = "mean", na.rm = TRUE)

library(ggplot2)
qplot(x=interval, y=steps, data = weekdaySummary, facets = weekday ~ ., geom = "path", xlab = "Intervals", ylab = "Steps")
```

As might be expected, the activity patterns are different. There is higher activity on weekdays before 8am, perhaps indicating a morning commute to work. There is a similar spike of activity between 8am and 9am throughout the week - perhaps indicating a morning exercise routine. There is a lower level of activity during the rest of the daylight hours on weekdays, perhaps indicating a sedentary job.